\documentclass{ximera}

\input{../preamble.tex}


\outcome{Understand linear density and its connection to mass.}
\outcome{Calculate the mass of an objection with varying density.}
\outcome{Understand work and how it is computed.}
\outcome{Calculate work when force is varying.}
\outcome{Know when to integrate a cross-section to solve a physics problem.}
\outcome{Calculate work when distance is varying.}


\title[Dig-In:]{Phyical applications}

\begin{document}
\begin{abstract}
We see applications of integration.
\end{abstract}
\maketitle

\section{Density}

Given an object which can be modeled as one dimensional (like a wire),
its \dfn{linear density} is a measure of its mass per unit length.

\begin{question}
  If a wire has linear density $5 \unit{g}/\unit{m}$, how many grams
  will $4 \unit{m}$ of this wire be?
  \[
  \text{Mass} = \answer{20} \unit{g}
  \]
\end{question}

Sometimes the linear density of an object can vary from one part of
the object to another.  Then we will have to use calculus to find the
total mass.

\begin{question}
  If the linear density of a rod of length $5\unit{m}$ is given by
  $p(x) = 1+3x^2 \unit{g}/\unit{m}$, where $x$ is the distance from
  one end of the rod, then what is total mass of the rod?
  \[
  \textrm{Mass} = \answer{130} \unit{g}
  \]
  \begin{hint}
    We can think of breaking the rod into infinitesimal pieces of
    length $\d x$.  At position $x$, the mass of the infinitesimal
    piece will be $(1+3x^2)\d x$.  So we just have to sum all of these
    up from $x=0$ to $x=5$.
  \end{hint}
  \begin{hint}
    \begin{align*}
      \text{Mass} &= \int_0^1 1+3x^2 \d x\\
      &=\eval{x+x^3}_0^5\\
      &=130
    \end{align*}
  \end{hint}
\end{question}

\section{Work}

On the other hand, \textit{work} is the scientific term used to
describe the action of a force which moves an object. When a constant
force $F$ is applied to move an object a distance $d$, the amount of
work performed is $W=F\cdot d$.

The SI unit of force is the newton, ($\unit{N} = \unit{kg}\cdot
\unit{m}/\unit{s}^2$), and the SI unit of distance is a meter
($\unit{m}$). The fundamental unit of work is one newton-meter, or a
joule ($\unit{J}$). That is, applying a force of one newton for one
meter performs one joule of work. In imperial units (as often used in
the United States), force is measured in pounds ($\unit{lb}$) and distance is
measured in feet ($\unit{ft}$, hence work is measured in foot-pounds.

\begin{warning} The concepts of \dfn{mass} and \dfn{weight} are
  closely related, yet different, concepts. The mass $m$ of an object
  is a quantitative measure of that object's resistance to
  acceleration. The weight $w$ of an object is a measurement of the
  force applied to the object by the acceleration of gravity $g$.

  Since the two measurements are proportional, $w=m\cdot g$, they are
  often used interchangeably in everyday conversation. However, when
  computing \textit{work}, one must be careful to note which is being
  referred to. When mass is given, it must be multiplied by the
  acceleration of gravity to reference the related force.  We will
  approximate standard gravity, on Earth, as
  $9.8\frac{\unit{m}}{\unit{s}^2}$ or as $32
  \frac{\unit{ft}}{\unit{s}^2}$
\end{warning}

When force is constant, the measurement of work is straightforward.

\begin{question}
Lifting a $200\unit{lb}$ object $5 \unit{ft}$ performs $\answer{1000}$
foot-pounds of work.
\end{question}

In the real world, most problems are not so simple.  Either the amount
of force changes as you move the object, or the distance each part of
the object has to move is variable.  We will need to use calculus to
calculate work in these kinds of situations.

In this course, we will either
\begin{itemize}
  \item \textbf{accumulate large forces over infinitesimal distances}, or we
    will
  \item \textbf{accumulate infinitesimal forces over large distances}.
\end{itemize}
Let's see an example of each type of situation:

\subsection{Accumulating large forces over infinitesimal distances}


\begin{example}
  How much work is performed pulling a $60\unit{m}$ climbing rope up a
  cliff face, where the rope has a mass of $66\unit{g}/\unit{m}$?
  \begin{explanation}
    Let $y$ be the distance we have pulled the rope.  The force $F(y)$
    on the rope is always ``large'' (read: not infinitesimal), but the
    force is changing as we pull the rope up.  If we think of pulling
    the rope up only an infinitesimal amount $\d y$, then the force
    will be constant over that infinitesimal distance, so the
    infinitesimal amount of work done by moving the rope through $\d
    y$ at height $y$ is $\d W = F(y) \d y$.  Integrating these
    infinitesimal amounts of work from $y=0$ to $y=60$ should yield
    the total amount of work done.	
    \begin{question}
      In this example, the mass of the rope remaining after $y$ feet have been pulled up is
      \[
      \mathrm{mass}(y) = \answer{(0.066)(60-y)} \unit{kg}
      \]
      \begin{hint}
	There are $60-y$ feet of rope, at a density of $66$ g/m, so
        the total mass of the rope is $(60-y)66$ g.  Converted to kg,
        this is $(60-y)(0.066)$ kg.
      \end{hint}
    \end{question}
    \begin{question}
      The force due to gravity on the rope after $y$ feet have been pulled up is
      \[
      F(y) = \answer{(0.066)(60-y)(9.8)} \unit{N} 
      \]
      \begin{hint}
	To get the force due to gravity from the mass, we just
        multiply the mass by standard gravity, $g = 9.8
        \frac{\unit{m}}{\unit{s}^2}$.  Since we have already written
        our mass in $\unit{kg}$, we obtain $F(y) = (0.066)(60-y)(9.8)
        \unit{N}$.
      \end{hint}
    \end{question}
   
    \begin{question}
      The total work done is
      \[
      \int_0^{60} F(y) \d y = \answer{1164.24} \unit{J}
      \]
      
      \begin{hint}
	\begin{align*}
	  \int_0^{60} F(y) \d y &= \int_0^{60}(0.066)(60-y)(9.8) \d y \\
	  &= (0.066)(9.8) \eval{60y - \frac{y^2}{2}}_0^{60}\\
	  &=(0.066)(9.8)(1800)\\
	  &= 1164.24
	\end{align*}
      \end{hint}
    \end{question}
    
    By comparison, consider the work done in lifting the entire rope
    $60$ meters. The rope weights $60\times 0.066 \times 9.8 =
    38.808\unit{N}$, so the work applying this force for 60 meters is
    $60\times 38.808 = 2,328.48 \unit{J}$. This is exactly twice the
    work calculated before, we leave it to the young mathematician to
    understand why.
  \end{explanation}
\end{example}

\subsection{Accumulating infinitesimal forces over large distances}


\begin{example}
  A cylindrical storage tank with a radius of $10 \unit{ft}$ and a
  height of $30\unit{ft}$ is filled with water, which weighs
  approximately $62.4 \unit{lb}/\unit{ft}^3$. Compute the amount of
  work performed by pumping the water up to a point $5$ feet above the
  top of the tank.
  \begin{explanation}
    If we were lifting the entire tank $35$ feet, there would be no
    need for calculus, and we could treat this as a simple
    multiplication problem.  Unfortunately, different parts of the
    water are traveling different distances.  The water at the top of
    the tank has a much shorter distance to travel than the water at
    the bottom of the tank, for instance.
	
    Our approach will be to look at infinitesimal slabs of water.  All
    of the water in such a slab will have to travel the same distance
    to the top of the tank.  The force due to gravity on this slab
    will be infinitesimal because the volume of the water is
    infinitesimal.
    
    Let $y$ be the depth of the slab below the top of the cylinder,
    and $\d y$ be its width.

    \begin{image}
      \begin{tikzpicture}
        \draw[penColor,very thick] (0,2) ellipse (2 and .7);
        \draw[very thick,penColor!20!background] (2,-2) arc (0:180:2 and .7);% top half of ellipse
        \draw[very thick,penColor] (-2,-2) arc (180:360:2 and .7);% bottom half of ellipse
      
        \draw[very thick,penColor!20!background] (2,0) arc (0:180:2 and .7);% top half of ellipse
        \draw[very thick,penColor] (-2,0) arc (180:360:2 and .7);% bottom half of ellipse
        \draw[very thick,penColor!20!background] (2,0.1) arc (0:180:2 and .7);% top half of ellipse
        \draw[very thick,penColor] (-2,0.1) arc (180:360:2 and .7);% bottom half of ellipse
        
        \draw[decoration={brace,mirror, raise=.1cm},decorate,thin] (-2,2)--(-2,0);
        
        \node [above,penColor] at (-2.4,0.7) {$y$};
        \node [above,penColor] at (2.2,-0.3) {$\d y$};
        
        \draw[penColor, very thick] (2,2) -- (2,-2);
        \draw[penColor, very thick] (-2,2) -- (-2,-2);
      \end{tikzpicture}
    \end{image}
    The total work done is
    \[
    \textrm{Work} = \answer{3744000 \pi} \text{foot-pounds}
    \]
    \begin{hint}
      Each slab has a volume of $\pi r^2 \d y = 100\pi \d y$
    \end{hint}
    \begin{hint}
      Thus the weight of each slab is $62.4(100\pi)\d y = 6240\pi \d
      y$.  Note that this is already a force, not a mass measurement,
      so we do not have to convert it further.
    \end{hint}
    \begin{hint}
      Each slab has to move a distance of $y+5$, so the total work
      done by moving each slab is $6240 \pi (y+5) \d y$.
    \end{hint}
    \begin{hint}
      Thus the total work done by pumping all of the water to the top is
      \begin{align*}
	\text{Work} &= \int_0^{30} 6240 \pi (y+5) \d y\\
	&= 6240 \pi \eval{\frac{1}{2}y^2+5y}_0^{30}\\
	&=  3744000 \pi
      \end{align*}
    \end{hint}
  \end{explanation}
\end{example}



\begin{example}
  \begin{warning}
    We expect you to memorize Hooke's law, and be able to use it in similar work problems.
  \end{warning}
  
  \textbf{Hooke's Law} states that the force required to compress or
  stretch a spring $x$ units from its natural length (its un-stretched
  length) is proportional to $x$; that is, this force is $F(x) = kx$
  for some constant $k$.

  For example, if a a force of $1\unit{N}$ stretches a given spring
  $2\unit{cm}$, then a force of $5\unit{N}$ will stretch the spring
  $10\unit{cm}$. Converting the distances to meters, we have that
  stretching this spring $0.02\unit{m}$ requires a force of $F(0.02) =
  k(0.02) = 1\unit{N}$, hence $k = 1/0.02 = 50\unit{N}/\unit{m}$.

  Say a force of $20\unit{lb}$ stretches a spring from its natural
  length of $7$ inches to a length of $12$ inches. How much work was
  performed in stretching the spring to this length?
  \[
  \textrm{Work} = \answer{\frac{50}{12}} \textrm{ft--lb}
  \]
  
  \begin{hint}
    This is a ``Large forces over infinitesimal distances'' problem.
  \end{hint}
  
  \begin{hint}
    Let $x$ be the amount we have stretched the spring beyond its
    natural length in inches, and let $F(x)$ be the force exerted by
    the spring at this distance in pounds.  We have that $F(0)=0$,
    $F(5) = 20$, and we know that it is linear in the distance by
    Hooke's law.  So we must have $F(x) = 4x$.
  \end{hint}
  
  \begin{hint}
    So the work done by moving the spring from $x$ to $x+\d x$ is $4x
    \d x$.  We need to accumulate these infinitesimal forces from
    $x=0$ to $x=5$.
  \end{hint}
  
  \begin{hint}
    \begin{align*}
      \textrm{Work} &= \int_0^5 4x \d x\\
      &= \eval{2x^2}_0^5\\
      &=50
    \end{align*}
  \end{hint}
  
  \begin{hint}
    We need the answer to be in foot-pounds not inch-pounds, so we convert
    inches to feet to obtain $\frac{50}{12}$ foot-pounds.
  \end{hint}
  
\end{example}

\end{document}
