\documentclass{ximera}

\input{../preamble.tex}


\outcome{Identify a differential equation.}
\outcome{Verify a solution to a differential equation.}
\outcome{Compute a general solution to a differential equation via integration.}
\outcome{Solve initial value problems.}
\outcome{Determine the order of a differential equation.}


\title[Dig-In:]{Differential equations}

\begin{document}
\begin{abstract}
  Differential equations show you relationships between rates of
  functions.
\end{abstract}
\maketitle

A \textit{differential equation}\index{differential equation} is
simply an equation with a derivative in it. Here is an example:
\[
a\cdot f''(x) + b\cdot f'(x) + c\cdot f(x) = g(x). 
\]
\begin{question}
  What is a differential equation?
  \begin{multipleChoice}
    \choice{An equation that you take the derivative of.}
    \choice[correct]{An equation that relates the rate of a function to other values.}
    \choice{It is a formula for the slope of a tangent line at a given point.}  
  \end{multipleChoice}
\end{question}

When a mathematician solves a differential equation, they are finding
\textit{functions} satisfying the equation.
\begin{question}
  Which of the following functions solve the differential equation
  \[
  f^{(4)}(x) = f(x)?
  \]
  \begin{hint}
    Remember, $f^{(4)}$ is the fourth derivative of $f$.
  \end{hint}
  \begin{selectAll}
    \choice[correct]{$f(x) = \sin(x)$}
    \choice{$f(x) = x^2$}
    \choice[correct]{$f(x) = e^x$}
    \choice[correct]{$f(x) = e^{-x}$}
    \choice{$f(x) = \tan(x)$}
  \end{selectAll}
  \begin{feedback}
    It turns out that the complete solution to this differential equation
    is $c_1\sin(x)+c_2\cos(x)+c_3e^x+c_4e^{-x}$.  In other words, every
    solution to this differential equation can be written in this form.
    You should check that these are all solutions (for example $f(x) =
    \sin(x)+3\cos(x)-7e^x+\pi e^{-x}$ is a solution).  Proving that these
    are \textbf{all} of the solutions is beyond the scope of this course.
  \end{feedback}
\end{question}


The differential equation above is an example of a fourth \dfn{order}
differential equation, because the highest derivative in the equation
is a ``fourth'' derivative. In general the highest derivative in a
differential equation is the order.


Differential equations are one of the most practical objects of
mathematical study.  They appear constantly in every field of science
and engineering.  They are a powerful way to model many diverse
situations.



\section{Exponential growth and decay}

A function $f$ exhibits \dfn{exponential growth} if its growth rate is
proportional to its value. As a differential equation, this means
\[
f'(x) = k f(x)\qquad\text{for some constant of proportionality $k$.}
\]
We claim that this differential equation is solved by
\[
f(x) = Ae^{kx},
\]
where $A$ and $k$ are constants.  Check it out, if $f(x) = Ae^{kx}$,
then
\begin{align*}
f'(x) &= Ak e^{kx}\\
&= k\left(Ae^{kx} \right)\\
&= k f(x).
\end{align*}

Here are some examples of real-world phenomona that can be modeled by exponential
growth:
\begin{itemize}
  \item The rate that a population grows is proportional to the size
    of the population.
  \item The rate that a radioactive substance decays is proportional
    to the amount of the substance.
  \item The rate that an investment grows is proportional to the
    amount of the investment.
  \item The rate that a pathogen spreads is proportional to the number
    of people with the pathogen.
  \item The rate that glucose is processed in one's blood is
    proportional to the concentration of glucose in the blood.
\end{itemize}



\begin{example}
A culture of yeast starts with $100$ cells. After $160$ minutes, there
are $350$ cells. Assuming that the growth rate of the yeast is
proportional to the number of yeast cells present, estimate when the
culture will have $1000$ cells.

\begin{explanation}
Since the growth rate of the yeast is proportional to the number of
yeast cells present, we have the following differential equation
\[
p'(t) = k p(t)
\]
where $p(t)$ is the population of the yeast culture at time $t$ with
$t$ measured in minutes. We know that this differential equation is
solved by the function
\[
p(t) = A e^{kt}
\]
where $A$ and $k$ are yet to be determined constants. Since
\[
100 = p(0) = Ae^{k\cdot 0}
\]
we see that $A = 100$. So 
\[
p(t) = 100 e^{kt}.
\]
Now we must find $k$. Since we know that 
\[
\answer[given]{350} = p(160) = 100e^{k\cdot 160}
\]
we need to solve for $k$. Write
\begin{align*}
350 &= 100 e^{k\cdot 160}\\
3.5 &= e^{k\cdot 160}\\
\ln(3.5) &= k\cdot 160\\
\ln(3.5)/160 &= k. 
\end{align*}
Hence
\[
p(t) = 100 e^{t\ln(3.5)/160} = 100 \cdot 3.5^{t/160}.
\]
To find out when the culture has 1000 cells, write
\begin{align*}
1000 &= 100 \cdot 3.5^{t/160}\\
10 &= 3.5^{t/160}\\
\ln(10) &= \frac{t\ln(3.5)}{160}\\
\frac{160\ln(10)}{\ln(3.5)} &= t.
\end{align*}
From this we find that after approximately $294$ minutes, there are
around $1000$ yeast cells present.
\end{explanation}
\end{example}

It is worth seeing an example of exponential decay as well. Consider
this: Living tissue contains two types of carbon, a stable isotope
carbon-12 and a radioactive (unstable) isotope carbon-14. While an
organism is alive, the ratio of one isotope of carbon to the other is
always constant. When the organism dies, the ratio changes as the
radioactive isotope decays.  This is the basis of radiocarbon dating.


\begin{example}
The half-life of carbon-14 (the time it takes for half of an amount of
carbon-14 to decay) is about 5730 years. Moreover, the rate of decay
of carbon-14 is proportional to the amount of carbon-14.

If we find a bone with $1/70$th of the amount of carbon-14 we would
expect to find in a living organism, approximately how old is the
bone?

\begin{explanation}
Since the rate of decay of carbon-14 is proportional to the amount of
carbon-14 present, we can model this situation with the differential
equation
\[
f'(t) = k f(t).
\]
We know that this differential equation is solved by the function
defined by
\[
f(t) = A e^{kt}
\]
where $A$ and $k$ are yet to be determined constants. Since the
half-life of carbon-14 is about $5730$ years we write
\[
\frac{1}{2} = e^{k 5730}.
\]
Solving this equation for $k$, gives
\[
k = \frac{-\ln(2)}{5730}.
\]
Since we currently have $1/70$th of the original amount of carbon-14
we write
\[
\answer[given]{\frac{1}{70}} = 1\cdot e^{\frac{-\ln(2)t}{5730}}.
\]
Solving this equation for $t$, we find $t \approx -35121$. This means
that the bone is approximately $35121$ years old.
\end{explanation}
\end{example}



\section{Modeling with differential equations}



\begin{question}
  Imagine that a glass of water has initial temperature $5^\circ
  \unit{C}$, and that the ambient temperature is $22^\circ \unit{C}$.
  The water will warm up over time.  Assume that the rate of change in
  the temperature of the water is directly proportional to the
  difference between the current water temperature and the ambient
  temperature.  Which of the following differential equations must be
  satisfied by the function $H(t)$ which measures the temperature of
  the water with respect to time?
  \begin{multipleChoice}
    \choice{$y' = 5+\frac{y}{22}$}
    \choice[correct]{$y' = k(22-y)$ for some $k>0$}
    \choice{$y' = k(y-22)$ for some $k>0$}
    \choice{$y' = k(5-y)$ for some $k>0$}
    \choice{$y' = k(y-5)$ for some $k>0$}
  \end{multipleChoice}
  \begin{hint}
    This is just a straight translation job.  ``The rate of change in
    the temperature of the water'' is $y'$.  ``Directly proportional
    to'' means that it is equal to some constant (say $k$) times
    whatever it is proportional to.  ``The difference between the
    current water temperature and the ambient temperature" is either
    $22-y$ or $y-22$, since $y$ is the temperature of the water and
    $22$ is the ambient temperature.  Think about which we should
    choose before looking at the next hint.  Will it be $y'=k(22-y)$
    or $y'=k(y-22)$ where $k>0$?
  \end{hint}
  \begin{hint}
    Since the temperature of the water is increasing over time, we
    want $y'>0$.  Since the temperature will be increasing (and it is
    reasonable to assume it never surpasses the ambient temperature!)
    $22-y$ is positive.  So we can conclude that $y' = k(22-y)$ for
    some $k>0$.
  \end{hint}
  
  \begin{feedback}
    The differential equation does not involve the number $5$.  If we
    wanted to incorporate that piece of data into our model we could
    ask ``Which solution(s) to this differential equation satisfy
    $H(0) = 5$?''  This is known as an \textbf{initial value problem}.  %%%
  \end{feedback}
\end{question}


Sometimes the rates in question are constant.



\begin{question}
  One can approximate the force of gravity as constant near the Earth.
  So the acceleration of a falling object is a constant $g>0$.  If
  $h(t)$ is the height of an object at time $t$, which differential
  equation must $h$ satisfy?
  \begin{multipleChoice}
    \choice{$y' = g$}
    \choice{$y=-g$}
    \choice[correct]{$y''=-g$}
    \choice{$y'' = g$}
    \choice{$y'' = -gy$}
  \end{multipleChoice}
  
  \begin{hint}
    The acceleration of an object is the second derivative of its
    position, so the differential equation should say the second
    derivative ($y''$) is constant.  Should it be a positive of
    negative constant?
  \end{hint}
  \begin{hint}
    A falling object will fall quicker and quicker, so the second
    derivative of its height should be negative.  Thus $y''=g$ is the
    correct answer.
  \end{hint}	
\end{question}

We have already seen, and solved, a particular kind of differential
equation in this course.  Namely a solution $F$ to the differential
equation $y' = f(x)$ is just an antiderivative of $f$!  We know the
``general solution'' of this differential equation is just $F(x)+C$,
as long as the domain of $f$ is an interval.  We can use this idea to
solve differential equations of the form $y^{(n)} = f(x)$, by just
repeatedly integrating.

\begin{example}
  Let's find the general solution to the differential equation $y'' =
  x$, and then find the particular solution which passes through the
  points $(0,1)$ and $(1,2)$.
  
  Since $y'' = x$, we know that $y' = \int x \d x$. 
  
  Thus $y' = \frac{1}{2} x^2 + C_1$ for some constant $C_1$.  
  
  Now this further implies that $y = \int \frac{1}{2} x^2 + C_1 \d x$, so 
  we must have that $y = \frac{1}{6}x^3+C_1x+C_2$, for some constant $C_2$.
  
  This is the general solution of the differential equation.  We have shown that every solution is of the form $y = \frac{1}{6}x^3+C_1x+C_2$.
  
  Now to find the particular solution we are interested in, we can just solve a system of equations.
  
  We find that the only solution to this differential equation passing through $(0,1)$ and $(1,2)$ is 
  
  \[
  y = \answer{\frac{1}{6}x^3+\frac{5}{6}x+1}
  \]
  
  \begin{hint}
    Substituting $(0,1)$ into the equation yields $C_2 = 1$, and then
    substituting $(1,2)$ into the equation one can solve for $C_1$ to
    find $C_1 = \frac{5}{6}$
  \end{hint}
  
  
\end{example}



\end{document}
