\documentclass{ximera}

\input{../preamble.tex}


\outcome{Find the equations for planes.}
\outcome{Give the equation for a sphere or ball.}

\title[Dig-In:]{Two and three dimensional geometry}

\begin{document}
\begin{abstract}
  blug
\end{abstract}
\maketitle

\subsection{Lines}

Parametric formulas for lines that start at a point when $t=0$ and
progress in a direction given by a vector can be described in
\textit{any} dimension as
\[
\l(t) = \mathrm{point} + t\cdot \vec{vector}
\]
\begin{question}
  Give a line that starts at the point $(1,2,3)$ give a parametric
  formula such that that ....
\end{question}

If no vector is given, we can find it by subtracting.... 


In three dimensions we have three coordinates axes, the $x$, $y$, and
$z$ axis:

\begin{image}
\begin{tikzpicture}
\begin{axis}[
  view={35}{15},
  axis lines=center,
  width=15cm,height=15cm,
  xtick={-10,-5,5,10},ytick={-10,-5,5,10},ztick={-10,-5,5,10},
  minor tick={-12,-11,...,12},
  xmin=-11,xmax=11,ymin=-11,ymax=11,zmin=-11,zmax=11,
 % xlabel={$x$},ylabel={$y$},zlabel={$z$},
]


\node [above right] at (axis cs:10,0,0) {$x$};
\node [above right] at (axis cs:0,10,0) {$y$};
\node [above right] at (axis cs:0,0,10) {$z$};


\end{axis}
\end{tikzpicture}
\end{image}

Together, these form three coordinate planes (The $(x,y)$-plane, the
$(x,z)$-plane, and the $(y,z)$-plane).

\begin{question}
	The $yz$-plane corresponds to which of the following equations?
	
	\begin{multipleChoice}
		\choice[correct]{$x=0$}
		\choice{$y=0$}
		\choice{$z=0$}
	\end{multipleChoice}
	
	\begin{hint}
	  Every point on the $yz$ axis has $x=0$, so this is the answer.
	\end{hint}
\end{question}

\begin{question}
	Which of the following most accurately describes the solution
        set of $y=2$ in $\R^3$?
	
	\begin{multipleChoice}
		\choice{A plane parallel to the $xy$ plane}
		\choice[correct]{A plane parallel to the $xz$ plane}
		\choice{A plane parallel to the $yz$ plane}
	\end{multipleChoice}

\begin{hint}
	$y=2$ consists of all those points where $y=2$, but $x$ and
  $z$ are allowed to be anything.  This corresponds to a plane which
  is parallel to the $xz$ plane ($y=0$).
\end{hint}
	
\end{question}

\begin{question}
	The equation $(x-1)^2+y^2+(z+2)^2 = 4$ has a solution set in
        $\R^3$ which is a sphere.  What is the center and
        radius of this sphere?
	
\[
\text{Radius} = \answer{2}
\]

\[
\text{Center} = \left(\answer{1},\answer{0}, \answer{-2} \right)
\]

\begin{hint}
	$(x-1)^2+y^2+(z+2)^2$ is the square of the distance from
  $(1,0,-2)$ to $(x,y,z)$.  If the square of the distance is $4$, then
  the distance is $2$.  Since the solution set of this equation is all
  points which are a distance of $2$ away from $(1,0,-2)$, then this
  is a sphere of radius $2$ centered at $(1,0,-2)$
\end{hint}
\end{question}

\end{document}
